---
title: "Building_Dataframes_Consolidated"
output: html_document
date: "2025-04-14"
---
# Setup 
```{r}
library(dplyr)
library(readxl)
library(readr)
library(tidyr)
```

# Load relevant dataframes 
```{r}
taxa_filtered <- read_xlsx(here("Data/Modified/Collapsed_Species_Code_DFs/taxa_filtered.xlsx"))
taxa <- read.csv(here("Data/Unmodified/taxa.csv"))
quad_freq <- read.csv(here("Data/Unmodified/Quadrat_Frequency.csv"))
quad_freq_filtered <- read_xlsx(here("Data/Modified/Collapsed_Species_Code_DFs/quad_freq_filtered.xlsx"))

#Build Simple Taxa
simple_taxa <- taxa %>%
  group_by(Genus, Species) %>%
  summarise(Code1 = first(Species_Code),
            Genus = first(Genus),
            Species = first(Species)) %>%
  ungroup()

#Make a genus_species column on both taxa and simple_taxa 
taxa <- taxa %>% mutate(Genus_Species = paste(Genus, Species, sep="_"))
taxa_filtered <- taxa_filtered %>% mutate(Genus_Species = paste(Genus, Species, sep="_"))
simple_taxa <- simple_taxa %>% mutate(Genus_Species = paste(Genus, Species, sep="_"))
```

# Creation of Collapsed Codes (combining subspecies and varieties, so that individuals are grouped by genus+species)
```{r}

  #make a genus_species column on both taxa and simple_taxa 
      taxa <- taxa %>% mutate(Genus_Species = paste(Genus, Species, sep="_"))
      taxa_filtered <- taxa_filtered %>% mutate(Genus_Species = paste(Genus, Species, sep="_"))
      simple_taxa <- simple_taxa %>% mutate(Genus_Species = paste(Genus, Species, sep="_"))
      
  #add the Code1 column to taxa 
      taxa_filtered <- taxa_filtered %>%
        left_join(simple_taxa %>% select(Genus_Species, Code1),
                  by = "Genus_Species")

  #join code1 to quad_freq using old code 
      quad_freq <- quad_freq %>%
        left_join(taxa %>% select(Species_Code, Code1),
                  by = "Species_Code")
  #make columns to help with joining 
      quad_freq <- quad_freq %>% mutate(Quad_Num = paste(Transect, Quadrat, sep = "_"))
      quad_freq <- quad_freq %>% mutate(Plot_Year = paste(Plot, Sample_Year, sep="_")) %>%
        group_by(Plot) %>%
        mutate(Quad_Total = n_distinct(Quad_Num))
      
      
      
#Checking to see which species are only keyed out to genus level 
genus_only <- taxa_filtered %>%
  filter(Genus != "" & !is.na(Genus) & Species == "" | is.na(Species)) %>%
  select(Species_Code, Vascular_Code, Genus, Species)
genus_only

#Checking for duplicates in vascular codes 
    duplicates <- taxa_filtered %>%
      filter(Vascular_Code != "" & !is.na(Vascular_Code)) %>%
      group_by(Genus, Species) %>%
      filter(n() > 1) %>%
      ungroup()
    duplicates
    
    duplicates_true <- duplicates %>% filter(Is_Name_Current %in% c("TRUE"))
    duplicates_false <- duplicates %>% filter(Is_Name_Current %in% c("FALSE"))
```      

# Checking for duplicate sample visits within the same year and removing them 
```{r}
  #see what plots have multiple visits within the same year 
      multiple_visits <- quad_freq %>%
        group_by(Vascular_Code, Plot, Sample_Year) %>%
        filter(n_distinct(Sample_Date) > 1)  %>%
        ungroup() %>%
        select(Vascular_Code, Plot, Sample_Year, Sample_Date)
      multiple_visits <- multiple_visits %>% distinct()
      
#Other duplicates - multiple quadrat repeats (ie two transects ran for quality assurance calibration)
katm_2009_02_032 <- vascular_only %>% filter(Plot %in% c("KATM_2009_02_032"))
duplicate_quads2 <- katm_2009_02_032 %>%
  group_by(Quad_Num, Code1) %>%
  filter(n() > 1) 
lacl_2007_02_002 <- vascular_only %>% filter(Plot %in% c("LACL_2007_02_002"))
duplicate_quads2 <- lacl_2007_02_002 %>%
  group_by(Quad_Num, Code1) %>%
  filter(n() > 1) 
      
  #remove duplicate sampling events 
      quad_freq_filtered <- quad_freq %>%
        filter(!(Plot == "LACL_2007_02_002" & Sample_Date == "2022-07-17") &
                 !(Plot == "LACL_2007_02_006" & Sample_Date == "2007-08-13"))
    
```
      
      
# Making the quadrat frequency dataframes with collapsed codes and corrected sample years 
```{r}
  #create relevant columns that aid with grouping and summarizing 
    quad_freq_filtered <- quad_freq_filtered %>% mutate(Quad_Num = paste(Transect, Quadrat, sep = "_"))
    quad_freq_filtered <- quad_freq_filtered %>% mutate(Plot_Year = paste(Plot, Sample_Year, sep="_")) %>%
        group_by(Plot) %>%
        mutate(Quad_Total = n_distinct(Quad_Num))
    
  #filter by classification 
      vascular_only_filtered <- quad_freq_filtered %>% filter(Vascular_Code %in% c("Vascular"))
      lichen_only_filtered <- quad_freq_filtered %>% filter(Vascular_Code %in% c("Lichen"))
      nonvasc_only_filtered <- quad_freq_filtered %>% filter(Vascular_Code %in% c("Nonvascular"))
    
  #create quadrat frequency dataframes 
      vasc_df <- vascular_only_filtered %>% 
        group_by(Plot_Year, Code1) %>% 
        summarise(Species_Quad_Count =n_distinct(Quad_Num)) %>% 
        ungroup() 
      quad_abundance_df_vasc <- vasc_df %>%
        pivot_wider(names_from = Code1, values_from = Species_Quad_Count, values_fill = 0)       
        
      lichen_df <- lichen_only_filtered %>% 
        group_by(Plot_Year, Code1) %>% 
        summarise(Species_Quad_Count =n_distinct(Quad_Num)) %>% 
        ungroup() 
      quad_abundance_df_lichen <- lichen_df %>%
        pivot_wider(names_from = Code1, values_from = Species_Quad_Count, values_fill = 0)
      
      nonvasc_df <- nonvasc_only_filtered %>% 
        group_by(Plot_Year, Code1) %>% 
        summarise(Species_Quad_Count =n_distinct(Quad_Num)) %>% 
        ungroup() 
      quad_abundance_df_nonvasc <- nonvasc_df %>%
        pivot_wider(names_from = Code1, values_from = Species_Quad_Count, values_fill = 0)
      
      
  #transform the quadrat frequency dataframes into presence-absence example code: 
  presence_absence_df <- quad_abundance_df %>%
      mutate(across(-Plot_Year, ~ifelse(. > 1, 1, .)))
  presence_absence_df <- presence_absence_df %>%
  separate(Plot_Year, into = c("Park", "EstYear", "Elevation_Band", "PlotID", "Sample_Year"), sep = "_", remove = FALSE)
```

# Creation of Summary Tables - Used to Help Balance Plot Visits 

## Vascular 
```{r}
summary_table_vasc <- quad_abundance_df_vasc  %>%
  group_by(Plot) %>%
  summarise(
    Number_Visits = n_distinct(Sample_Year), 
    Visit_Years = paste(sort(unique(Sample_Year)), collapse = ", "),
    Vegetation_Class = unique(Vegetation_Class))
summary_table_vasc <- summary_table_vasc %>% 
  arrange(Vegetation_Class, Number_Visits)
summary_table_vasc <- summary_table_vasc %>%
  left_join(vasc_abundance_df %>% select(Plot, Viereck.2, Viereck.3),
            by = "Plot")
summary_table_vasc <- summary_table_vasc %>% distinct()
```

## Lichen
```{r}
summary_table_lichen <- quad_abundance_df_lichen  %>%
  group_by(Plot) %>%
  summarise(
    Number_Visits = n_distinct(Sample_Year), 
    Visit_Years = paste(sort(unique(Sample_Year)), collapse = ", "),
    Vegetation_Class = unique(Vegetation_Class))
summary_table_lichen <- summary_table_lichens %>% 
  arrange(Vegetation_Class, Number_Visits)
summary_table_lichen <- summary_table_lichens %>%
  left_join(lichen_abundance_df %>% select(Plot, Viereck.1, Viereck.2, Viereck.3),
            by = "Plot")
summary_table_lichen <- summary_table_lichens %>% distinct()
```

## Nonvascular 
```{r}
summary_table_nonvasc <- quad_abundance_df_nonvasc  %>%
  group_by(Plot) %>%
  summarise(
    Number_Visits = n_distinct(Sample_Year), 
    Visit_Years = paste(sort(unique(Sample_Year)), collapse = ", "),
    Vegetation_Class = unique(Vegetation_Class))
summary_table_nonvasc <- summary_table_lichens %>% 
  arrange(Vegetation_Class, Number_Visits)
summary_table_nonvasc <- summary_table_nonvasc %>%
  left_join(nonvasc_abundance_df %>% select(Plot, Viereck.1, Viereck.2, Viereck.3),
            by = "Plot")
summary_table_nonvasc <- summary_table_nonvasc %>% distinct()
```


# Balancing dataframes for PERMANOVA 

```{r}
#hand select the years for plots with more than the decided number of visits 
filter_criteria <- list(
  Plot1 = c(Selected_Year, Selected_Year, Selected_Year),
  Plot2 = c(Selected_Year, Selected_Year, Selected_Year),
  Plot3 = c(Selected_Year, Selected_Year, Selected_Year),
  Plot4	= c(Selected_Year, Selected_Year, Selected_Year))

#next identify plots with three plot visits from your vegetation class (or whatever the number you've decided is)
selected_plots <- summary_table_of_choice %>%
  filter(Vegetation_Class == "Desired Vegetation Class", Number_Visits == 3) %>%
  pull(Plot)

#combine plots with hand selected plots from filter criteria 
selected_plots <- union(selected_plots, names(filter_criteria))

#subset main presence absence dataframe based on both conditions 
#verify that the hand selected plots were correctly pulled 
subset_hand_selected <- quad_abundance_dataframe_of_choice %>%
  filter(
    Plot %in% names(filter_criteria) & 
      Sample_Year %in% unlist(filter_criteria[Plot]))

balanced_df <- quad_abundance_dataframe_of_choice %>%
  filter(
    (Plot %in% selected_plots) | 
      (Plot %in% names(filter_criteria) &
         Sample_Year %in% unlist(filter_criteria[Plot])))
```


# Creation of Species Richness Dataframes 
```{r}
  #Calculate species richness by classification 
    #Lichen 
        lichen_sp_richness <- lichen_only_filtered %>%
        group_by(Plot, Sample_Year) %>% 
        summarize(Species_Richness = n_distinct(Code1), .groups = "keep") 
        lichen_sp_richness <- lichen_sp_richness %>% mutate(Plot_Year = paste(Plot, Sample_Year, sep="_"))
      
    #Vascular 
      vasc_sp_richness <- vascular_only_filtered %>%
        group_by(Plot, Sample_Year) %>% 
        summarize(Species_Richness = n_distinct(Code1), .groups = "keep") 
        vasc_sp_richness <- vasc_sp_richness %>% mutate(Plot_Year = paste(Plot, Sample_Year, sep="_"))
      
    #Nonvascular 
      nonvasc_sp_richness <- nonvasc_only_filtered %>%
        group_by(Plot, Sample_Year) %>% 
        summarize(Species_Richness = n_distinct(Code1), .groups = "keep") 
        nonvasc_sp_richness <- nonvasc_sp_richness %>% mutate(Plot_Year = paste(Plot, Sample_Year, sep="_"))
      
  #Combine them into one total table
    #rename columns for clarity 
        vasc_sp_richness <- vasc_sp_richness %>% rename(Vasc_Species_Richness = Species_Richness)
        lichen_sp_richness <- lichen_sp_richness %>% rename(Lichen_Species_Richness = Species_Richness)
        nonvasc_sp_richness <- nonvasc_sp_richness %>% rename(Nonvasc_Species_Richness = Species_Richness)
        
    #combine them 
        combined_df <- vasc_sp_richness %>%
          left_join(lichen_sp_richness, by = c("Plot", "Sample_Year"))
        combined_df <- combined_df %>%
          left_join(nonvasc_sp_richness, by = c("Plot", "Sample_Year"))
        combined_df <- combined_df %>%
          left_join(viereck %>% select(Plot, Sample_Year, Vegetation_Class, Viereck.3), by = c("Plot", "Sample_Year"))
        
    #remove nonessential columns from joins 
        str(combined_df)  
        combined_df <- combined_df[, -4]
        combined_df <- combined_df[, -5]
        combined_df <- combined_df[, -6]
        
    #add viereck class as a column 
        viereck <- read.csv(here("Data/Unmodified/Viereck_Classes.csv"))
        combined_df <- combined_df %>%
          left_join(viereck %>% select(Plot, Sample_Year, Sample_Date), by = c("Plot", "Sample_Year"))
```


# Creation of ENV files (variable files for running PERMANOVA)

```{r}
        env <- df_of_choice[,c(1:6)] #exclude quadrat frequency data, include variables like year, viereck class, plot, etc. 
        
    #add visit column
        env <- env %>%
          arrange(Plot, Sample_Year) %>%
          group_by(Plot) %>%
          mutate(Visit = paste0("visit_", row_number())) %>%
          ungroup() 
    #add species richness column
        env <- env %>%
          left_join(sp_richness_df_of_choice %>% select(Plot_Year, Species_Richness), by = c("Plot_Year"))
    #adding canopy cover 
        env <- env %>%
          left_join(canopy_cover %>% select(Plot_Year, Percent_Cover), by = c("Plot_Year"))
        env <- env %>% distinct()
```


Random Code for Quick Reference from Dataframe_Management script
```{r}



```







































